---
title: "Supplemental Material"
author: 
- name: Patricia Hernandex
  email: hernandez.p@wustl.edu
  affiliation: Washington University in St. Louis, School of Medicine, St. Louis, MO, USA
- name: Nick Borcherding
  email: borcherding.n@wustl.edu
  affiliation: Washington University in St. Louis, School of Medicine, St. Louis, MO, USA
output: word
---

# Supplemental material

## Loading Library

This will load the libraries necessary to run the analysis. If the libraries/packages are not installed, this code will install them before loading them.

```{r message = FALSE, warning = FALSE}
required_libraries <- c(
  "dplyr", "data.table", "tidyr", "janitor", "stringr", "venn",
  "lubridate", "readxl", "openxlsx", "tidyverse", "hrbrthemes",
  "proustr", "VennDiagram", "ComplexUpset"
)

# Function to check, install if needed, and load libraries
load_libraries <- function(libraries) {
  for (lib in libraries) {
    if (!requireNamespace(lib, quietly = TRUE)) {
      install.packages(lib, dependencies = TRUE)
    }
    library(lib, character.only = TRUE)
  }
}

# Execute the function
load_libraries(required_libraries)
```

## Loading Functions

These are custom functions written for the analysis to assist in loading and plotting the data. More information can be found in the github repository under the R folder.

```{r message = FALSE, warning = FALSE}
source("./R/load_data.R")
source("./R/utils.R")
source("./R/calculate_sum.R")
source("./R/generate_venn_diagram.R")
source("./R/generate_upset_plot.R")

```

## Extra functions

Functions to be used for supplemental tables

```{r}
# Asterix
asterix<- function(list){lapply(list, function(x) {gsub("_", "*", x)})} #Replace _ with *

# Transform list into dataframe
list_to_df <- function(list){data.frame(lapply(list, function(x) {
  x <- unlist(x)
  length(x) <- max(lengths(list))
  return(x)
}))}


table_maker <- function(df, filename = "output.docx", add_new_page = FALSE) {
  # Define the directory "./outputs/tables"
  directory <- "./outputs/tables"
  
  # Create the directory if it doesn't exist
  if (!dir.exists(directory)) {
    dir.create(directory, recursive = TRUE)
    cat("Directory created at:", directory, "\n")
  }
  
  # Ensure the filename ends with ".docx"
  if (!grepl("\\.docx$", filename)) {
    stop("Filename must end with '.docx'")
  }
  
  # Define the full output path
  output_path <- file.path(directory, filename)
  
  # Check if the document exists
  if (file.exists(output_path)) {
    # Load the existing document
    doc <- read_docx(path = output_path)
  } else {
    # Create a new document
    doc <- read_docx()
  }
  
  # Create a flextable from the dataframe
  ft <- flextable(df) %>%
    theme_box() %>%
    autofit()
  
  # Add a new page if required
  if (add_new_page) {
    doc <- doc %>% body_add_break()
  }
  
  # Add the flextable to the document
  doc <- doc %>% body_add_flextable(ft)
  
  # Save the document
  print(doc, target = output_path)
  
  # Return the path of the saved file
  cat("Table added to Word document at:", output_path, "\n")
  return(output_path)
}

```

### Figure 1 - Class I

```{r message=FALSE, warning=FALSE}
sab <- process_sab_data(file = "./data/SAB_Class_I_Lot_14_20221212.xlsx", c(1,2)) 
pra <- process_pra_data(file = "./data/PRA_Class_I_Lot_020_20221212.xlsx", 1) 
mix <- process_mix_data(file = "./data/LSM12NC23_024_01.xlsx", 1) 
explex <- process_explex_data(file = "./data/ExPlex_classI_II.xlsx", 1)

#Remove non-type-I from Mix
mix <- mix[grepl("A_|B_|C_", mix$values),]

b <- list(
  SAB = unique(sab$Molecular.Typing),
  PRA = unique(pra$values),
  Mix = unique(mix$values),
  ExPlex = unique(explex$Molecular.Typing)
)

b <- asterix(b)

```

```{r}
unlistb <- list_to_df(b)
#table_maker(unlistb,filename = "STable1_ClassI.docx")
```

#### Only intersections

```{r}
#PRA, SAB, Mix
intersection <- Reduce(intersect, unlistb[, c("SAB", "PRA", "Mix")])
print(intersection)
```

```{r}
# ExPlex
noexplex <- c(unlistb$SAB,unlistb$PRA,unlistb$Mix)
ex <-unlistb$ExPlex[!unlistb$ExPlex %in% noexplex]
```

### Figure 1 - Class II

```{r message = FALSE, warning = FALSE}
sab <- process_sab_data(file = "./data/SAB_ClassII_Lot_15_20221212.xlsx", c(1:2))[-c(1:2),]
pra <- process_pra_data(file = "./data/PRA_Class_II_Lot_019_20221212.xlsx", 1)
mix <- process_mix_data(file = "./data/LSM12NC23_024_01.xlsx", 1) 
explex <- process_explex_data(file = "./data/ExPlex_classI_II.xlsx", 2)

#Remove type-I from Mix
mix <- mix[!grepl("A_|B_|C_", mix$values),]

b <- list(
  SAB = unique(sab$Molecular.Typing),
  PRA = unique(pra$values),
  Mix = unique(mix$values),
  ExPlex = unique(explex$Molecular.Typing)
)
b <- asterix(b)
```

```{r}
unlistb <- list_to_df(b)
#table_maker(unlistb,filename = "STable1_ClassI.docx")
```

#### Only intersections

```{r}
#PRA, SAB, Mix
intersection <- Reduce(intersect, unlistb[, c("SAB", "PRA", "Mix")])
print(intersection)
```

```{r}
# ExPlex
noexplex <- c(unlistb$SAB,unlistb$PRA,unlistb$Mix)
ex <-unlistb$ExPlex[!unlistb$ExPlex %in% noexplex]
print(ex)
```

#### SAB and ExPlex mutual eplets (class II)

```{r}
#SAB and ExPlex
intersection2 <- Reduce(intersect, unlistb[, c("SAB","ExPlex")])
print(intersection2)
```

## Figure 2, Class II

```{r}
df <- process_hla_registry(file = "./data/HMM-CL02-0620-CAT-00.xlsx", 1)
sab <- process_sab_data(file = "./data/SAB_ClassII_Lot_15_20221212.xlsx", c(1:2))[-c(1:2),]
pra <- process_pra_data(file = "./data/PRA_Class_II_Lot_019_20221212.xlsx", 1)
mix <- process_mix_data(file = "./data/LSM12NC23_024_01.xlsx", 1)
explex <- process_explex_data(file = "./data/ExPlex_classI_II.xlsx", 2)

#SAB and ExPlex
intersection2 <- intersect(sab$Molecular.Typing,explex$Molecular.Typing)
print(intersection2)

```

```{r}
# Intersection SAB and ExPlex - eplets 
filtered_df <- df[grep(intersection2, df$Alleles), ]
```
